---
title: ''
author: ''
date: ''
output: 
  html_document:
    df_print: paged
    keep_md: true
    toc: TRUE
    toc_float: TRUE
    theme: flatly 
    code_download: TRUE
  word_document:
    reference_docx: style.1.docx
---

```{r setup, include = FALSE}

  proj <- rprojroot::find_rstudio_root_file()
  knitr::opts_knit$set(root.dir = proj)

  knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
                        fig.width  = 6 * 1.67, fig.height = 6)
  
  load(paste0(proj, "/data/er_data.RData"))
  source(paste0(proj, "/scripts/functions.R"))
  
  library(descr)
  library(tidyverse)
  library(viridis)
  library(flextable)
  library(patchwork)
  library(testthat)
  library(knitr)

  # See QUERY (for others) and NOTE (to self) tags throughout. 

```

# Objectives

The overall objective is to relate nurse staffing to patient outcomes, and in particular the **patient experience time** (PET), **time to triage** (TTT) and the **wait to be seen time** (WBS). Staffing is reflected in terms of the actual, rostered, and required staff levels for the same period, for both RNs and HCAs. 

The key analytical challenge is that the data are inherently multilevel. The unit of observation in the study is the patient-visit, meaning that there are multiple visits per patient and so the patient-visits are not independent. Further, staff levels are measured for both 24 hour periods and shifts, so that patients and patient-visits are also clustered within these time periods. Things are further complicated by the fact that patient-visits can and do span across multiple periods (shifts or 24 hour periods). I suspect that we will have to make some simplifying selections and assumptions, such as only considering first visits, or assuming that patient-visits are only linked to the time-period they enter into. 

The other challenge is to account for the effects of nurse staffing **relative** to actual needs for that time-period. This is reflected in some degree in the "required" staff levels, but might also need to consider the specific case-mix of patients in the ER for that period, as well as the trolley count, day of the week, etc. 

# Units of observation

There are a total of `r nrow(data)` observations (rows) in the dataset, which describes `r length(unique(data$patient_code))` individual patients. 

Patients can have multiple visits to the ER (with the visit order recorded in `data$visit`). Their arrival times to the ER are logged and range from `r min(data$ed_arrive_date_time, na.rm = TRUE)` to `r max(data$ed_arrive_date_time, na.rm = TRUE)`.

```{r}

  # Data check: 
  expect_equal(nrow(data), 121966)
  expect_equal(
    min(data$ed_arrive_date_time, na.rm = TRUE), 
    as.POSIXct("2018-01-01 00:12:00 GMT")
    )
  expect_equal(
    max(data$ed_arrive_date_time, na.rm = TRUE), 
    as.POSIXct("2020-04-01 00:04:00 BST")
  )

```


```{r}

  # Data check: 
  # Do first_visit times match up with the corresponding ed_arrive_date_time?

  df <- filter(data, visit == 1) %>% 
  select(first_visit, ed_arrive_date_time) %>%
  map_df(as.numeric)
  # ggplot(df, aes(first_visit, ed_arrive_date_time)) +
  #   geom_point() +
  #   coord_fixed()
  
  expect_equal(cor(df$first_visit, df$ed_arrive_date_time, use = "complete"), 1) 
  # Perfect match

```

The rows of the dataset can then be uniquely identified by two columns: `patient_code` and `visit` (or `ed_arrive_date_time`). 

```{r}

  # Data check
  # Are rows uniquely identified by patient_code and visit or arrival time?

  expect_equal(nrow(distinct(data, patient_code, visit)), nrow(data))
  # expect_equal(nrow(distinct(data, patient_code, ed_arrive_date_time)),
  #              nrow(data))
  
  # df <- group_by(data, patient_code) %>%
  #   summarise(u_v = length(unique(visit)), 
  #             u_t = length(unique(ed_arrive_date_time)), 
  #             equal = u_v == u_t)
  # 
  # tar <- filter(df, equal == FALSE)
  # View(filter(data, patient_code %in% tar[[1]]))
  # Patients with funky dates here

```

Some patients had many visits (up to `r max(data$visit)`), though most only had one (`r round(prop.table(table(data$visit))[1]*100, 1)`%), and `r round(sum(prop.table(table(data$visit))[1:5])*100, 1)`% had 5 or fewer. 

Figure: Distribution of patient visits

```{r}

  g1 <- group_by(data, patient_code) %>%
    summarise(max_visits = max(visit)) %>%
    ggplot(aes(x = max_visits)) +
      geom_bar(fill = viridis(1, begin = 0.5)) +
      xlab("Number of total visits") +
      ggtitle("All patients")

  g2 <- group_by(data, patient_code) %>%
      summarise(max_visits = max(visit)) %>%
      ggplot(aes(x = max_visits)) +
        geom_bar(fill = viridis(1, begin = 0.5)) +
        xlab("Number of total visits") +
        ylim(0, 50) +
        ggtitle("Patients where < 50 had that many visits", 
                subtitle = "(presented separately so we can see the counts)")
  
  g1 / g2
  
```

# Patient-visit level characteristics

Some patient-visits are characterized by a variable called `new_attendances`. A "new attendance" is distinguished from "returns", which was any attendance that wasn't a "scheduled return" or an "other return". Of the `r nrow(data)` patient visits, `r round(prop.table(table(data$new_attendances))[2]*100, 2)`% were new attendances (i.e. not returns), and 100% of first visits were also new attendances (**QUERY**: is this a function of how new attendances and returns were defined based on dates in the database, or is this also by design?)   

```{r}

  data$new_attendances <- factor(
    data$new_attendances, labels = c("Return", "New attendence")
    )

  # The code below is just a scratchpad to see how well things match up with 
  # respect to the new attendances vs return visits

  # Note on new attendances from Gearoid (2021-12-20)
  # For 'New attendances', first, any patients on visit 1 who had their
  # discharge destination marked as returns (numbers 44, 45, or 46), if they had
  # a new arrival within 42 days of the ed dept date of visit 1, visit 2 was
  # marked as a scheduled return. If patients had a visit 2 within 28 days of a
  # visit 1, and the discharge destination of visit 1 was not marked a return,
  # visit 2 was marked as 'other return'. Overall returns were then any visit
  # which occurred within 28 days of that patients last visit. New attendances
  # were then any attendance that wasn't a 'scheduled return' or an 'other
  # return'.

  # How many first visits (in the database) are returns?
  # table(data$new_attendances[data$visit == 1]) # All of them

  # Scheduled returns in 42 days
  # with(data, table(new_attendances, scheduled_return_42_days))
  # 24 scheduled returns, all marked as returns. 
  
  # Any return in 28 days?
  # with(data, table(new_attendances, overall_returns_28_days))
  # 23746 returns w/in 28 days, all classed as returns except 1 (QUERY)
  
  # Data check:
  # See the the time difference between the 2nd and 1 visits match to the 28 day
  # return flag
  
  # df <- filter(data, visit == 2) %>%
  #   mutate(
  #     visit_diff = factor(
  #       as.numeric(
  #         difftime(
  #           ed_arrive_date_time, first_visit, units = "days")
  #       ) < 28, 
  #       labels = c("28+ days", "<28 days")
  #     )
  #   )
  
  # with(df, table(visit_diff, overall_returns_28_days))
  # Most of these are correct (194 + 54 misclassified)
  
  # QUERY: should we bother "correcting" these?

  # NOTE: This might need more checking, but overall looks close enough to not
  # bother with for now in order to move things forward. 

```


## Key Outcomes

The main outcomes, measured at the patient-visit level, are based on 4 key date-times:

Arrival to ER -> Triage -> Seen -> Leave ER

We want to plot these times to see if they make sense. 

First, let's see if they are all in the right order. In the plot below, I've taken every patient-visit, selected the 4 events of interest (Arrival, Triage, Seen, Discharge), and ordered them by their recorded date-times.

Figure: Observed order of key time-points

```{r}

  times <- c("ED Arrival", "ED Depart", "Seen", "Triage")

  select(data, patient_code, visit, contains("date_time")) %>% 
    mutate(patient_code = reorder(factor(patient_code), ed_arrive_date_time)) %>%
    gather(event, date, -patient_code, -visit) %>%
    mutate(event = factor(event, labels = times)) %>%
    mutate(event = factor(event, levels = times[c(1, 4, 3, 2)])) %>%
    arrange(patient_code, date) %>%
    group_by(patient_code, visit) %>%
    mutate(order = row_number()) %>%
    ungroup() %>%
    group_by(event) %>%
    mutate(mean_order = mean(order)) %>%
    ungroup() %>%
    mutate(event = reorder(event, mean_order)) %>% 
  ggplot(aes(x = order, y = event, color = event)) +
    geom_jitter(size = 0.5, alpha = 0.1) +
    scale_color_viridis(guide = FALSE, discrete = TRUE) +
    theme_minimal() +
    theme(panel.grid = element_blank())
```

In the plot above, each point is patient-visit-time (there are so many they look like solid blocks on color in some squares). What it shows is that ED arrival is always first (except for 6 barely visible dots there in the 2nd position on the x axis). Triage is *almost* always second (almost all the dots for triage times are in the order = 2 square, but a few are found in 3rd or 4th order); and Seen and Discharge mostly fall into the 3rd or 4th positions, but not always. 

It is important to note the percentage of missing values for each key event date-time:

Figure: Percent missing for each key event

```{r}

  select(data, contains("date_time")) %>%
  map_df(is.na) %>% 
  gather(event, missing) %>%
  mutate(event = factor(event, labels = times)) %>%
  mutate(event = factor(event, levels = times[c(1, 4, 3, 2)])) %>%
  ggplot(aes(x = event, fill = missing)) +
    geom_bar(position = "fill") +
    ylab("Percent missing") +
    scale_fill_viridis("Missing?", discrete = TRUE, begin = 0.1, end = 0.5) +
  xlab("")

```

Table: Percent missing for key events (corresponding to the plot above)

```{r}

  select(data, contains("date_time")) %>%
    map_df(function(x)table(is.na(x))) %>%
    rename(no_miss = "FALSE", miss = "TRUE") %>%
    mutate(var = times, pct_miss = round(100 * miss / (miss + no_miss), 2)) %>%
    select(var, everything()) %>%
    kable()

```

So virtually everyone has an arrival and departure time, while the vast majority have a triage time, and most have a seen time. 

```{r}

  # Flag patients for whom all the dates are in order

  data <- data %>%
    mutate(
      correct_order =
        ed_arrive_date_time < triage_start_date_time & 
        triage_start_date_time < seen_date_time & 
        seen_date_time < ed_depart_date_time
    ) 
  # Note: this variable is missing for anyone without all 4 times, so let's 
  # expand to other groups with 3 of 4 times, but still in the right order. 

  data$correct_order <- ifelse(
    (data$ed_arrive_date_time < data$seen_date_time & 
     data$seen_date_time == data$triage_start_date_time &
     data$triage_start_date_time < data$ed_depart_date_time), 
    "A/S-T/D Correct", 
    data$correct_order
  )
  
  data$correct_order <- ifelse(
    is.na(data$correct_order) & 
    (data$ed_arrive_date_time < data$triage_start_date_time & 
     data$triage_start_date_time < data$ed_depart_date_time), 
    "A/T/D Correct", 
    data$correct_order
  )
  
  data$correct_order <- ifelse(
    is.na(data$correct_order) & 
    (data$ed_arrive_date_time < data$seen_date_time & 
     data$seen_date_time < data$ed_depart_date_time), 
    "A/S/D Correct", 
    data$correct_order
  )
  
  data$correct_order <- ifelse(
    is.na(data$correct_order) & 
    (data$ed_arrive_date_time < data$ed_depart_date_time), 
    "A/D Correct", 
    data$correct_order
  )
  
  data$correct_order <- factor(
    data$correct_order, 
    labels = c("A/D Correct", "A/S-T/D Correct", "A/S/D Correct", 
               "A/T/D Correct", "Incorrect", "A/T/S/D Correct")
  )
  
  data <- full_join(
    data, 
    group_by(data, correct_order) %>% summarise(n = n()), 
    by = "correct_order"
  ) %>% 
    mutate(correct_order = reorder(correct_order, n)) %>%
    select(-n)
  
  # filter(data, is.na(correct_order)) %>%
  #   select(contains("date_time")) %>%
  #   View() 
  # This just leave 48 patient-visits with an arrival time only, 3 with only
  # arrival and triage times, and 1 with a seen time before arrival and
  # discharge.
    

  # with(data, table(visit, correct_order)) 
  # These are not clearly related to visits. 
  
  # filter(data, correct_order == "Incorrect") %>%
  #   select(contains("date_time")) %>%
  #   View()
  
```

Now we consider overall patterns of missingness and the degree to which key date-times were in the correct order across all the patient-visits. There were `r table(data$correct_order)["A/T/S/D Correct"]` patient-visits where there were no missing date-times and they were all in the correct order (labeled **A/T/S/D Correct**; `r round(table(data$correct_order)["A/T/S/D Correct"]/nrow(data)*100, 2)`% of the total database); and another `r table(data$correct_order)["A/T/D Correct"]` patient-visits where there were no missing date-times for arrival, triage, and discharge, and they were all in the correct order (labeled **A/T/D Correct**; `r round(table(data$correct_order)["A/T/D Correct"]/nrow(data)*100, 2)`% of the total database). The counts for the rest are given in the plot below. 

```{r}

  ggplot(data, aes(x = correct_order)) +
    geom_bar(fill = viridis(1, begin = 0.5)) +
    ylab("Number of patient-visits")

```

These match up with "not triaged" status. 

Figure: Relationship between triage status and date-time patterns (proportional and absolute counts)

```{r}

  g1 <- ggplot(data, aes(x = correct_order, fill = triage_category)) +
    geom_bar(position = "fill") +
    scale_fill_viridis(guide = FALSE, discrete = TRUE) +
    ylab("Percent") +
    coord_flip() +
    xlab("")

  g2 <- ggplot(data, aes(x = correct_order, fill = triage_category)) +
    geom_bar() +
    scale_fill_viridis("", discrete = TRUE) +
    ylab("Count") +
    coord_flip() +
    theme(legend.position = "top")
  
  g1 / g2

```

Figure: Timing for for patient-visits where key date-times were out of order

```{r}

  times <- c("ED Arrival", "ED Depart", "Seen", "Triage")

  filter(data, correct_order == "Incorrect") %>%
    select(patient_code, visit, contains("date_time")) %>%
    mutate(patient_code = reorder(factor(patient_code),
                                  ed_arrive_date_time)) %>%
    gather(which, time, -patient_code, -visit) %>%
    mutate(which = factor(which, labels = times)) %>%
    mutate(which = factor(which, levels = times[c(1, 4, 3, 2)])) %>%
  ggplot(aes(x = time, y = patient_code, color = which, group = patient_code)) +
    geom_point() +
    geom_line(alpha = 0.5) +
    scale_color_viridis("", discrete = TRUE, end = 0.9) +
    xlab("") +
    ylab("Patient") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid = element_blank())

```

You can see that for most of these patient-visits the date-times all sit on top of each other (at this scale). We will explore the actual time differences below. 

Here is what these same data look like for the first 100 patient-visits that were first visits and that have all four key date-times in the expected order. 

```{r}

  times <- c("ED Arrival", "ED Depart", "Seen", "Triage")

  filter(data, correct_order == "A/T/S/D Correct" & visit == 1) %>%
    slice_head(n = 100) %>%
    select(patient_code, contains("date_time")) %>%
    mutate(patient_code = reorder(factor(patient_code), ed_arrive_date_time)) %>%
    gather(which, time, -patient_code) %>%
    mutate(which = factor(which, labels = times)) %>%
    mutate(which = factor(which, levels = times[c(1, 4, 3, 2)])) %>%
  ggplot(aes(x = time, y = patient_code, color = which, group = patient_code)) +
    geom_point() +
    geom_line(alpha = 0.5) +
    scale_color_viridis("", discrete = TRUE, end = 0.9) +
    xlab("") +
    ylab("Patient") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid = element_blank())

```

Table: Distributions of key timespans

```{r}

  descript <- data %>%
   select(arrive_to_triage_hours, arrive_to_seen_hours, triage_to_seen_hours, 
          seen_to_discharge_hours)

  data_frame(Variable = name.1(descript),
             Observations   = n.miss(descript),
             "Mean SD" = summary.1(descript),
             "Median [IQR]" = med.iqr(descript),
             "(Min, Max)" = min.max(descript)) %>% knitr::kable()

```

The data above include everyone, including the small numbers of patients with out of order date-times, hence the negative value for arrive-to-seen time. 

Figure: Length of key timespans across all patient-visits (first 24 hours)

```{r}

  select(data, ends_with("_hours")) %>% 
    gather(timespan, hours) %>%
    mutate(
      timespan = 
        factor(timespan, levels = levels(factor(timespan))[c(3, 5, 2, 4, 1)])
      ) %>%
    ggplot(aes(x = hours, fill = timespan)) +
      geom_density(color = "black") +
      scale_fill_viridis(guide = FALSE, discrete = TRUE, end = 0.8) +
      facet_wrap(~timespan, ncol = 1, scales = "free_y") +
      xlim(0, 24)

```

## Other patient-visit level variables

```{r}

  descript <- data %>%
   select(age_at_admission, gender, mode_of_arrival, triage_category)

  data_frame(Variable = name.1(descript),
             Observations   = n.miss(descript),
             "Mean SD" = summary.1(descript),
             "Median [IQR]" = med.iqr(descript),
             "(Min, Max)" = min.max(descript)) %>% knitr::kable()

```

Referral and discharge will probably need some further categorizing to be useful. 

Figure: Referral types

```{r}

  data <- left_join(
    data, 
    group_by(data, referral_description) %>% summarise(n = n()), 
    by = "referral_description"
  ) %>%
  mutate(referral_description = reorder(referral_description, n)) %>%
  select(-n)

  ggplot(data, aes(x = referral_description, fill = referral_description)) +
    geom_bar() + 
    coord_flip() +
    scale_fill_viridis(guide = FALSE, discrete = TRUE, direction = -1)

```

Figure: Discharge destination

```{r}

  data <- left_join(
    data, 
    group_by(data, discharge_destination) %>% summarise(n = n()), 
    by = "discharge_destination"
  ) %>%
  mutate(discharge_destination = reorder(discharge_destination, n)) %>%
  select(-n)

  ggplot(data, aes(x = discharge_destination, fill = discharge_destination)) +
    geom_bar() + 
    coord_flip() +
    scale_fill_viridis(guide = FALSE, discrete = TRUE, direction = -1)

```

Does "left before seen" match up with expected date-times?

```{r}

  df <- filter(data, discharge_destination == "18 : Left before Seen")

  table(df$correct_order)
  
```

So some people who were discharged without being seen also have date-times for being seen (but only about 10% of them). 


# ER level factors

As noted above, each row in the dataset is a patient-visit, and this clustering of visits within patients needs to be accounted for in any analysis (i.e. patient-visits are not independent of one another). However, we also have another level of dependence, which is that patient-visits are also clustered over time, so that concurrent patient visits are more likely to share an environment characterized by a variety of time-varying factors including nurse staffing levels and trolly numbers.

## Shift vs 24 hour period

One way to capture is dependency is to consider patient-visits as nested within shifts. As such there is a `shift_code` in the dataset. These codes denote day and night shifts, and then a number that is generally in order with `ed_arrive_date_time`, but not always, as the plots below illustrate. Overall, there were `r length(unique(data$shift_code))` shifts in the dataset, of which `r length(unique(data$shift_code[grepl("D", data$shift_code)]))` were day shifts (one fewer than the number of night shifts). 


```{r}

  data %>%
  group_by(shift_number) %>%
  summarise(start = min(ed_arrive_date_time, na.rm = TRUE)) %>%
  ggplot(aes(start, as.numeric(shift_number))) +
    geom_point() +
    xlab("ED Arrival") +
    ylab("Shift number")

```

Zoom in where things fall our of order:

```{r}

  data %>%
  filter(as.numeric(shift_number) > 700) %>%
  group_by(shift_number) %>%
  summarise(start = min(ed_arrive_date_time, na.rm = TRUE)) %>%
  ggplot(aes(start, as.numeric(shift_number))) +
    geom_point() +
    xlab("ED Arrival") +
    ylab("Shift number")

```

**QUERY**: Is this discrepancy something that needs to be resolved?

Everything is in order for the 24 periods.

## Nursing staff levels

### Shift

Nursing levels are described in various ways and are consistently linked to the shift code.

```{r}

  # The key shift level variables are the various staffing level factors. The
  # first question to ask is whether these vary *within* shift, or if they
  # really are descriptors of an entire shift. If staffing levels are consistent
  # across shifts, then there should only be on unique value for each shift. So
  # here we plot the number of unique values to get a handle of whether this is
  # true, and how much variability there is if it isn't true.

  df <- data %>%
  select(shift_code, ends_with("_shift")) %>%
  gather(type, numbers, -shift_code) %>%
  group_by(shift_code, type) %>%
  summarise(unique_values = length(unique(numbers))) 

  expect_equal(length(unique(df$unique_values)), 1)

```

Figure: Distribution of nursing levels within shifts 

```{r}

  df <- data %>%
  select(shift_code, ends_with("_shift"), shift_dn) %>%
  distinct() %>%
  gather(type, numbers, -shift_code, -shift_dn) 

  ggplot(df, aes(x = numbers, fill = shift_dn)) +
    geom_histogram(data = filter(df, shift_dn == "day")) +
    geom_histogram(data = filter(df, shift_dn == "night"), alpha = 0.5,
                   color = "black") +
    facet_wrap(~type) +
    scale_fill_viridis("Day/Night", discrete = TRUE, end = 0.8) +
    theme(legend.position = "bottom")


```

### 24 hour periods

```{r}

  # Same as above for shift. 

  df <- data %>%
  select(x24hr_period_code, ends_with("_24hr")) %>%
  gather(type, numbers, -x24hr_period_code) %>%
  group_by(x24hr_period_code, type) %>%
  summarise(unique_values = length(unique(numbers))) 

  expect_equal(length(unique(df$unique_values)), 1) # All good

```

Figure: Distribution of nursing levels within 24 hour periods 

```{r}

  df <- data %>%
  select(x24hr_period_code, ends_with("_24hr")) %>%
  distinct() %>%
  gather(type, numbers, -x24hr_period_code) 

  ggplot(df, aes(x = numbers)) +
    geom_histogram(color = "black", fill = viridis(1, begin = 0.5)) +
    facet_wrap(~type)

```

## Trolley GAR

There is also information on the trolley count ([trolley GAR](https://www.hse.ie/eng/about/who/acute-hospitals-division/emergency-departments/)) that may help explain staffing levels relative to needs. The challenge of measuring staffing levels relative to needs is probably the crux of the analysis. 

```{r if_html, eval = knitr::opts_knit$get("rmarkdown.pandoc.to") == "html"}

# use flextable (works for MD files on github too)

```

```{r if_word, eval = !knitr::opts_knit$get("rmarkdown.pandoc.to") == "html"}

# use kable

```

```{r code_book}

# print(summarytools::dfSummary(data), method = "render")

```

```{r sysinfo}

  DescTools::SysInfo()

```

